\ab{Constraints on the quasi-normal mode frequencies of the LIGO-Virgo signals by making full use of gravitational-wave modeling}

### 2021-01-25 papers/intro_paper/intro_paper.tex
done with Sec. II

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo
Collaboration\sout{s}~\citep{Virgo} \ab{have} recently announced their catalogue of
gravitational-wave (GW) \sout{observations} \ab{signals}~\citep{GWTC-2} from the first
half of the\sout{ir} third observing run (O3a)~\citep{O3reference}. Combined
with \sout{the confirmed observations from} the first and second observing-run \ab{catalogues}~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford,
Washington and Livingston, Louisiana~\citep{aasi2015characterization},
and the Advanced Virgo detector in Cascina,
Italy~\citep{acernese2014advanced} have now detected \sout{more than} $50$ GW
events from the merger of compact objects like neutron stars and/or
black holes (BHs).\sout{, collectively called compact binary coalescences
(CBCs).} Alongside independent claims of
detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020},
\sout{this} \ab{these results} \sout{has} \ab{have} firmly established the field of GW astronomy, five years \sout{on
from} \ab{after} the first \sout{ever direct} detection of a GW \ab{passing through the} \sout{on} Earth, \ab{notably GW150914} 
\sout{on September 14,2015}~\citep{abbott2016observation}.

@@

\ab{\section{Waveform model and statistical strategy to measure quasi-normal modes}}

\comment{AB: In the following, I will not use much the strike-out, but instead I will 
bracket in color the text that I have changed.}

A GW signal from the (quasi-circular) coalescence of two BHs is
completely described in GR by a \sout{fixed set of} $15$ parameters,
$\bxigr$. These can be grouped into the \emph{intrinsic} parameters:
the masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the \ab{component 
objects in the} \sout{initial} binary; and the 
\emph{extrinsic} parameters: \ab{a reference time $t_c$ and phase 
$\phi_c$,} the sky position of the binary ($\alpha$,
$\delta$), the luminosity distance, $d_L$, and \ab{the binary's} orientation
described through the inclination of the binary $\iota$ and its
polarization $\psi$. \ab{We also introduce the total mass $M = m_1+m_2$, 
and the (dimensionless) symmetric mass ratio $\nu = m_1m_2/M^2$}.

\ab{Here, we focus on BHs with spins aligned or anti-aligned 
with the orbital angular momentum (henceforth, ``aligned-spin''). In this case,  
the GW signal depends on $11$ parameters. We denote the 
aligned-spin \ab{(dimensionless)} components as $\chi_{i} = |\vec{\bm{S}}_i|/m^2_i$, where $i=1,2$ for the two BHs.}

\ab{\subsection{Parameterized waveform model}}\label{sec:model}

As in \paperone, we use an IMR waveform model developed within the effective-one-body (EOB) 
formalism~\cite{}. \ab{However, whereas \paperone was limited to non-spinning multipolar waveforms, 
here we use as our baseline model the aligned-spin multipolar waveform model 
developed in Ref.~\citep{Cotesta:2018fcv}}. In addition to being
calibrated to NR simulations, this model also uses information from BH
perturbation theory for the merger and ringdown phases. \sout{Therefore this
waveform model provides a very accurate and faithful semi-analytic
description of the inspiral, merger and ringdown phases.} \ab{Henceforth we
will denote this model by $\SEOB$ for short}~\footnote{\ab{In the LIGO Algorithm Library (LAL), this 
waveform model is called {\tt SEOBNRv4HM}.}}.

@@

h_+(\iota,\varphi;t ) - i h_\times(\iota,\varphi;t) = \sum_{\ell, m} {}_{-\!2}Y_{\ell m}(\iota,\varphi)\, h_{\ell m}(t)\,,

@@

\comment{AB: there is a mismatch of notations for the parameters with respect to what was introduced above.}
where $\varphi$ is the azimuthal direction to the observer, ${}_{-\!2}Y_{\ell m}(\theta,\varphi)$ are the $-2$ spin-weighted spherical harmonics. The $\SEOB$ model we employ includes the $(\ell, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$ modes~\cite{Cotesta:2018fcv}. For each $(\ell, m)$, the inspiral-(plunge-)merger-ringdown $\SEOB$ waveform is schematically given by

@@

where $\theta(t)$ is the Heaviside step function, $h_{\ell m}^\mathrm{insp-plunge}$ represents the inspiral-plunge part of the waveform, whereas $h_{\ell m}^\mathrm{merger-RD}$ denotes the merger-ringdown \ab{waveform, which reads~\citep{Bohe:2016gbl,Cotesta:2018fcv}}

@@

% in terms of $c_{1,f}^{\ell   m},\ c_{2,f}^{\ell m},\ \sigma^\textrm{R}_{\ell m},\ |h_{\ell    m}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{\ell  m})|,\ \partial_t|h_{\ell    m}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{\ell m})|$ as

@@

In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the
complex frequencies $\sigma_{\ell m 0}$ are expressed in terms of the
final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the
latter are related to the BBH's component masses and spins through
NR--fitting-formulas obtained in
GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the
spirit of what was done in \paperone, we promote the QNM (complex)
frequencies to be free parameters of the model, while keeping the
inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed
to their GR values. \ab{More explicitly, we introduce a parameterized 
version of the $\SEOB$ model where the frequency and the
damping time of the ${\ell m 0}$ mode (i.e, $(f_{\ell m 0}, \tau
_{\ell m 0})$) is defined through the fractional deviations, $(\delta
f_{\ell m 0},\delta \tau_{\ell m 0})$, from the corresponding GR
values, which are obtained using NR fits~\cite{Taracchini:2013rva,Hofmann:2016yih}. Thus,} 
\begin{eqnarray}
f_{\ell m 0} &=& f_{\ell m 0}^{\text{GR}}\, (1 + \delta f_{\ell m 0})\,,\label{eq:nongr_freqs_a} \\ 
\tau _{\ell m 0} &=& \tau _{\ell m 0}^{\text{GR}}\, (1 + \delta \tau_{\ell m 0})\,. \label{eq:nongr_freqs_b}
\end{eqnarray}
\ab{We denote such a parameterized waveform model $\pSEOB$}~\footnote{\ab{This 
waveform model is called {\tt pSEOBNRv4HM} in LAL.}}.

\ab{As said}, the GR quantities $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
constructed using the same NR--fitting--formula \sout{prescriptiion} 
described above. We note that when leaving $\sigma_{\ell m}$ to vary
freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell
  m}(t)$ \sout{will} in general also differ from the GR prediction\ab{s}, since
those functions depend on the QNM complex frequencies, as can be seen
from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$ \ab{in Eqs.~(\ref{c1}), 
(\ref{c2}), and (\ref{d1}).}

\ab{\subsection{Bayesian parameter-estimation technique}}
\label{sec:method}

The parameterized model, $\pSEOB$, described above introduces an additional set of non-GR parameters, $\bxingr = (\delta f_{\ell m 0},\delta \tau_{\ell m 0})$, corresponding to each $(\ell,m)$ QNM present in the GR waveform model $\SEOB$. One then proceeds to use the Bayes theorem to obtain the \emph{posterior} probability distribution on $\blambda = \{\bxigr, \bxingr\}$, given a hypothesis $\mathcal{H}$:
%
\begin{equation}
P(\blambda | d, \mathcal{H}) = \frac{P(\blambda | \mathcal{H}) \, \mathcal{L}(d | \blambda, \mathcal{H})}{P(d|\mathcal{H})},
\label{eq:Bayes_theorem}
\end{equation}
%
where $P(\blambda | \mathcal{H})$ is the \emph{prior} probability distribution, and $\mathcal{L}(d | \blambda, \mathcal{H})$ is called the \emph{likelihood} function. The denominator is a normalization constant $P(d|\mathcal{H}) = \int P(\blambda | \mathcal{H}) \, \mathcal{L}(d | \blambda, \mathcal{H}) \, d\blambda$, called the marginal likelihood, or the \emph{evidence} of the hypothesis $\mathcal{H}$. In this case, our hypothesis $\mathcal{H}$ is that the data contains a GW signal that is described by the $\pSEOB$ waveform model $h(\blambda)$  and stationary Gaussian noise described by a power spectral density (PSD) $S_n(f)$. The likelihood function can consequently be defined as:
%
\begin{equation}
\mathcal{L}(d | \blambda, \mathcal{H}) \propto \exp\big[-\frac{1}{2} \langle d - h(\blambda) \, | \, d -h(\blambda) \rangle \big],
\label{eq:likelihood}
\end{equation}
%
where $\langle . | . \rangle$ is the following noise-weighted inner product:
%
\begin{equation}
\langle A | B \rangle = \int_{f_\mathrm{low}} ^{f_\mathrm{high}} df \frac{\tilde{A}^*(f)\tilde{B}(f) + \tilde{A}(f)\tilde{B}^*(f)}{S_n(f)}.
\label{eq:nwip}
\end{equation}
%
\ab{The quantity} $\tilde{A}(f)$ denotes the Fourier transform of $A(t)$ and the $^*$ indicates complex conjugation. The limits of integration ${f_\mathrm{low}}$ and ${f_\mathrm{high}}$ define the bandwidth of the sensitivity of the GW detector. We usually assume ${f_\mathrm{high}}$ to be the Nyquist frequency whereas${f_\mathrm{low}}$ is dictated by \ab{the performance of the 
GW detector at low-frequency and we assume XXX}. \comment{AB: It isn't just seismic noise. In fact, at low frequency, there are mysterious noises that people have had hard time 
to control.} \sout{the response of the gravitational detector to low-frequency seismic noise.} Owing to the large dimensionality of the parameter set $\blambda$, the posterior distribution $P(\blambda | d, \mathcal{H})$ in Eq.~(\ref{eq:Bayes_theorem}) is computed by stochastically sampling the parameter space using techniques such as Markov-chain Monte Carlo (MCMC)~\cite{} or Nested Sampling~\cite{}. For this paper, we use the \verb+LALInference+~\cite{} and \verb+Bilby+ codes~\cite{} that provide an implementation of the parallely tempered MCMC and Nested Sampling algorithms respectively, for computing the posterior distributions.

@@

For most of the results discussed in this paper, we restrict ourselves
to the $(\ell m) = (2,2)$ and/or $(3,3)$ modes. In those cases we
\sout{would} assume $\bxingr = \{\df{220},\dtau{220}\}$ and/or $
\{\df{330},\dtau{330}\}$, and all the other $(\ell m)$ modes to be
fixed at their GR predictions (i.e., $\delta f_{\ell m 0} = \delta
\tau_{\ell m 0} = 0$). This is because, for most of the high-mass BH 
events that we find most appropriate for this test, the LIGO-Virgo
observations are consistent with nearly--equal-mass face-on/off BBHs 
for which power in the subdominant modes is not enough to
attempt to measure more than one, or at most \comment{AB: it seems some wrods are missing 
to make the senetence meaningful} QNM frequencies, 
simultaneously.

\ab{Lastly}, throughout our analysis, we assume uniform priors on our non-GR QNM
parameters, $(\delta f_{\ell m 0},\delta \tau_{\ell m 0})$. However,
since the priors on $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
derived through NR--fits, from the corresponding priors on the initial
masses and spins, this leads to a non-trivial prior on the final
reconstructed frequency and damping time, $( f_{\ell m 0},\tau_{\ell m
  0})$. Also, given the definition of the damping time in
Sec.~\ref{sec:model}, we \sout{realize} \ab{note} that $\delta \tau_{\ell m 0} = -1$ leads
to the imaginary part of the frequency going to infinity. We avoid
this by restricting the minimum of the prior on $\delta \tau_{\ell m
  0}$ to be greater than $-1$.

>>>>

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo
Collaboration~\citep{Virgo} have recently announced their catalogue of
gravitational-wave (GW) signals from the first
half of the third observing run (O3a)~\cite{Abbott:2020niy}. Combined
with the first and second observing-run catalogues~\cite{LIGOScientific:2018mvr}, the Advanced LIGO detectors at Hanford,
Washington and Livingston, Louisiana~\citep{aasi2015characterization},
and the Advanced Virgo detector in Cascina,
Italy~\citep{acernese2014advanced} have now detected \sout{more than} $50$ GW
events from the merger of compact objects like neutron stars and/or
black holes (BHs).\sout{, collectively called compact binary coalescences
(CBCs).} Alongside independent claims of
detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020},
\sout{this} \ab{these results} \sout{has} \ab{have} firmly established the field of GW astronomy, five years \sout{on
from} \ab{after} the first \sout{ever direct} detection of a GW \ab{passing through the} \sout{on} Earth, \ab{notably GW150914} 
\sout{on September 14,2015}~\citep{abbott2016observation}.

@@

\section{Waveform model and statistical strategy to measure quasi-normal modes}

A GW signal from the (quasi-circular) coalescence of two BHs is
completely described in GR by a $15$ parameters,
$\bxigr$. These can be grouped into the \emph{intrinsic} parameters:
the masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the component 
objects in the binary; and the 
\emph{extrinsic} parameters: a reference time $t_c$ and phase 
$\phi_c$, the sky position of the binary ($\alpha$,
$\delta$), the luminosity distance, $d_L$, and the binary's orientation
described through the inclination of the binary $\iota$ and its
polarization $\psi$. We also introduce the total mass $M = m_1+m_2$, 
and the (dimensionless) symmetric mass ratio $\nu = m_1m_2/M^2$.

Here, we focus on BHs with spins aligned or anti-aligned 
with the orbital angular momentum (henceforth, ``aligned-spin''). In this case,  
the GW signal depends on $11$ parameters. We denote the 
aligned-spin (dimensionless) components as $\chi_{i} = |\vec{\bm{S}}_i|/m^2_i$, where $i=1,2$ for the two BHs.

\subsection{Parameterized waveform model}\label{sec:model}

As in \paperone, we use an IMR waveform model developed within the effective-one-body (EOB) 
formalism~\cite{Brito:2018rfr}. However, whereas \paperone was limited to non-spinning multipolar waveforms, 
here we use as our baseline model the aligned-spin multipolar waveform model 
developed in Ref.~\citep{Cotesta:2018fcv}. In addition to being
calibrated to NR simulations, this model also uses information from BH
perturbation theory for the merger and ringdown phases. Henceforth we
will denote this model by $\SEOB$ for short~\footnote{In the LIGO Algorithm Library (LAL), this 
waveform model is called {\tt SEOBNRv4HM}.}.

@@

h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{\ell, m} {}_{-\!2}Y_{\ell m}(\iota,\varphi_0)\, h_{\ell m}(t)\,,

@@

where $\varphi_0$ is the azimuthal direction to the observer \rb{Perhaps to clarify that $\varphi_0$ does not add another parameter we can say: "(note that without loss of generality we can take $\phi_c\equiv\varphi_0$)"}, ${}_{-\!2}Y_{\ell m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical harmonics. The $\SEOB$ model we employ includes the $(\ell, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$ modes~\cite{Cotesta:2018fcv}. For each $(\ell, m)$, the inspiral-(plunge-)merger-ringdown $\SEOB$ waveform is schematically given by

@@

where $\theta(t)$ is the Heaviside step function, $h_{\ell m}^\mathrm{insp-plunge}$ represents the inspiral-plunge part of the waveform, whereas $h_{\ell m}^\mathrm{merger-RD}$ denotes the merger-ringdown waveform, which reads~\citep{Bohe:2016gbl,Cotesta:2018fcv}

@@

@@

In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the
complex frequencies $\sigma_{\ell m 0}$ are expressed in terms of the
final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the
latter are related to the BBH's component masses and spins through
NR--fitting-formulas obtained in
GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the
spirit of what was done in \paperone, we promote the QNM (complex)
frequencies to be free parameters of the model, while keeping the
inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed
to their GR values. More explicitly, we introduce a parameterized 
version of the $\SEOB$ model where the frequency and the
damping time of the ${\ell m 0}$ mode (i.e, $(f_{\ell m 0}, \tau
_{\ell m 0})$) is defined through the fractional deviations, $(\delta
f_{\ell m 0},\delta \tau_{\ell m 0})$, from the corresponding GR
values, which are obtained using NR fits~\cite{Taracchini:2013rva,Hofmann:2016yih}. Thus,
\begin{eqnarray}
f_{\ell m 0} &=& f_{\ell m 0}^{\text{GR}}\, (1 + \delta f_{\ell m 0})\,,\label{eq:nongr_freqs_a} \\ 
\tau _{\ell m 0} &=& \tau _{\ell m 0}^{\text{GR}}\, (1 + \delta \tau_{\ell m 0})\,. \label{eq:nongr_freqs_b}
\end{eqnarray}
We denote such a parameterized waveform model $\pSEOB$~\footnote{This 
waveform model is called {\tt pSEOBNRv4HM} in LAL.}.

As said, the GR quantities $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
constructed using the same NR--fitting--formula 
described above. We note that when leaving $\sigma_{\ell m}$ to vary
freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell
  m}(t)$ in general also differ from the GR predictions, since
those functions depend on the QNM complex frequencies, as can be seen
from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$ in Eqs.~(\ref{c1}), 
(\ref{c2}), and (\ref{d1}).

\subsection{Bayesian parameter-estimation technique}
\label{sec:method}

The parameterized model, $\pSEOB$, described above introduces an additional set of non-GR parameters, $\bxingr = (\delta f_{\ell m 0},\delta \tau_{\ell m 0})$, corresponding to each $(\ell,m)$ QNM present in the GR waveform model $\SEOB$. One then proceeds to use the Bayes theorem to obtain the \emph{posterior} probability distribution on $\blambda = \{\bxigr, \bxingr\}$, given a hypothesis $\mathcal{H}$:
%
\begin{equation}
P(\blambda | d, \mathcal{H}) = \frac{P(\blambda | \mathcal{H}) \, \mathcal{L}(d | \blambda, \mathcal{H})}{P(d|\mathcal{H})},
\label{eq:Bayes_theorem}
\end{equation}
%
where $P(\blambda | \mathcal{H})$ is the \emph{prior} probability distribution, and $\mathcal{L}(d | \blambda, \mathcal{H})$ is called the \emph{likelihood} function. The denominator is a normalization constant $P(d|\mathcal{H}) = \int P(\blambda | \mathcal{H}) \, \mathcal{L}(d | \blambda, \mathcal{H}) \, d\blambda$, called the marginal likelihood, or the \emph{evidence} of the hypothesis $\mathcal{H}$. In this case, our hypothesis $\mathcal{H}$ is that the data contains a GW signal that is described by the $\pSEOB$ waveform model $h(\blambda)$  and stationary Gaussian noise described by a power spectral density (PSD) $S_n(f)$. The likelihood function can consequently be defined as:
%
\begin{equation}
\mathcal{L}(d | \blambda, \mathcal{H}) \propto \exp\big[-\frac{1}{2} \langle d - h(\blambda) \, | \, d -h(\blambda) \rangle \big],
\label{eq:likelihood}
\end{equation}
%
where $\langle . | . \rangle$ is the following noise-weighted inner product:
%
\begin{equation}
\langle A | B \rangle = \int_{f_\mathrm{low}} ^{f_\mathrm{high}} df \frac{\tilde{A}^*(f)\tilde{B}(f) + \tilde{A}(f)\tilde{B}^*(f)}{S_n(f)}.
\label{eq:nwip}
\end{equation}
%
The quantity $\tilde{A}(f)$ denotes the Fourier transform of $A(t)$ and the $^*$ indicates complex conjugation. The limits of integration ${f_\mathrm{low}}$ and ${f_\mathrm{high}}$ define the bandwidth of the sensitivity of the GW detector. We usually assume ${f_\mathrm{high}}$ to be the Nyquist frequency whereas ${f_\mathrm{low}}$ is dictated by the performance of the 
GW detector at low-frequency and we follow the choice made in the LIGO-Virgo analysis~\cite{LIGOScientific:2018mvr,Abbott:2020niy} \rb{Abhirup, can you check that indeed we followed the LIGO-Virgo analysis?}. Owing to the large dimensionality of the parameter set $\blambda$, the posterior distribution $P(\blambda | d, \mathcal{H})$ in Eq.~(\ref{eq:Bayes_theorem}) is computed by stochastically sampling the parameter space using techniques such as Markov-chain Monte Carlo (MCMC)~\cite{Metropolis:1953am,Hastings:1970aa} or Nested Sampling~\cite{Skilling:2006gxv}. For this paper, we use the \verb+LALInference+~\cite{Veitch:2014wba} and \verb+Bilby+ codes~\cite{Ashton:2018jfp,Smith:2019ucc} that provide an implementation of the parallely tempered MCMC and Nested Sampling algorithms respectively, for computing the posterior distributions.

@@

For most of the results discussed in this paper, we restrict ourselves
to the $(\ell m) = (2,2)$ and/or $(3,3)$ modes. In those cases we assume $\bxingr = \{\df{220},\dtau{220}\}$ and/or $
\{\df{330},\dtau{330}\}$, and all the other $(\ell m)$ modes to be
fixed at their GR predictions (i.e., $\delta f_{\ell m 0} = \delta
\tau_{\ell m 0} = 0$). This is because, for most of the high-mass BH 
events that we find most appropriate for this test, the LIGO-Virgo
observations are consistent with nearly--equal-mass face-on/off BBHs 
for which power in the subdominant modes is not enough to
attempt to measure more than one QNM complex frequency.

Lastly, throughout our analysis, we assume uniform priors on our non-GR QNM
parameters, $(\delta f_{\ell m 0},\delta \tau_{\ell m 0})$. We note that
since the priors on $( f_{\ell m 0}^{GR},\tau_{\ell m 0}^{GR})$ are
derived through NR--fits, from the corresponding priors on the initial
masses and spins, this leads to a non-trivial prior on the final
reconstructed frequency and damping time, $( f_{\ell m 0},\tau_{\ell m
  0})$. Also, given the definition of the damping time in
Sec.~\ref{sec:model}, we note that $\delta \tau_{\ell m 0} = -1$ leads
to the imaginary part of the QNM complex frequency going to infinity. We avoid
this by restricting the minimum of the prior on $\delta \tau_{\ell m
  0}$ to be greater than $-1$.
